---
title: 'Red Wine Quality Exploration '
output:
  html_document: default
  pdf_document: default
---

This report explored a dataset containing quality and attributes for approximately 1600 wines, and intented to answer: Which chemical properties influence the quality of red wines?


```{r, message=FALSE, warning=FALSE, packages, echo=FALSE}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(RColorBrewer)
library(scales)
library(memisc)
library(reshape)
```

### Data Summary
Let's look at a snapshot of the data and some summary statistics.  
This tidy data set contains 1,599 red wines with 11 variables on the chemical properties of the wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent). The dataset contains wine quality from 3 to 8.
```{r echo=FALSE}
# Load the Data
wine = read.csv('wineQualityReds.csv')
str(wine)
summary(wine)
```

Since we are most interested in the quality of wine, it will be helpful to look at the target variable.

```{r, Univariate_Plots}
wine$quality = factor(wine$quality)
ggplot(aes(x=quality), data = wine)+
  geom_histogram(stat="count", color = I('lightblue'),fill = I('lightblue'))
```

### Univariate Plots Section

11 attributes are analyzed in this report, including: 
 - fixed.acidity	
 - volatile.acidity  
 - citric.acid	
 - residual.sugar	
 - chlorides	
 - free.sulfur.dioxide	
 - total.sulfur.dioxide  
 - density	
 - pH	
 - sulphates	
 - alcohol

To first explore this data visually, I’ll draw up quick histograms of all 12 variables. The intention here is to see a quick distribution of the values.
 
Some other features that would also be interesting to explore but are not in this dataset.
 - year of the wine
 - origin of the wine
 - type of the wine, Cabernet Franc, Cabernet Sauvignon, Gamay, Grenache/Garnacha and Malbec etc.


```{r echo=FALSE}
par(mfrow=c(3,4)) 
hist(wine$fixed.acidity)
hist(wine$volatile.acidity)
hist(wine$citric.acid)
hist(wine$residual.sugar)
hist(wine$chlorides)
hist(wine$free.sulfur.dioxide)
hist(wine$total.sulfur.dioxide)
hist(wine$density)
hist(wine$pH)
hist(wine$sulphates)
hist(wine$alcohol)
```

### Univariate Analysis

- It appears that density and pH are normally distributed, with few outliers.
- Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol seem to be long-tailed.
- Qualitatively, residual sugar and chlorides have extreme outliers.
- Citric acid appeared to have a large number of zero values. I’m curious whether this is truly zero, or if it is a case of non-reporting.

### Bivariate Plots Section
```{r, Bivariate_Plots_0}
quality_groups = group_by(wine, quality)
wine.by_quality = summarise(quality_groups, 
          fixed.acidity_mean = mean(fixed.acidity),
          volatile.acidity_mean = mean(volatile.acidity),
          citric.acid_mean = mean(citric.acid),
          residual.sugar_mean = mean(residual.sugar),
          chlorides_mean = mean(chlorides),
          free.sulfur.dioxide_mean = mean(free.sulfur.dioxide),
          total.sulfur.dioxide_mean = mean(total.sulfur.dioxide),
          density_mean = mean(density),
          pH_mean = mean(pH),
          sulphates_mean = mean(sulphates),
          alcohol_mean = mean(alcohol),
          n = n())
head(wine.by_quality,10)
```


```{r, Bivariate_Plots}
WineQuality =  c(3,4,5,6,7,8)

c1 = cor.test(as.numeric(wine.by_quality$quality),wine.by_quality$fixed.acidity_mean)

p1 <- ggplot(aes(y=WineQuality, x=fixed.acidity_mean), data = wine.by_quality) +
  geom_line(size = 1)

p2 <- ggplot(aes(y= WineQuality, x=volatile.acidity_mean), data = wine.by_quality) +
  geom_line(size = 1)

p3 <- ggplot(aes(y= WineQuality, x=citric.acid_mean), data = wine.by_quality) +
  geom_line(size = 1)

p4 <- ggplot(aes(y= WineQuality, x=residual.sugar_mean), data = wine.by_quality) +
  geom_line(size = 1)

p5 <- ggplot(aes(y= WineQuality, x=free.sulfur.dioxide_mean), data = wine.by_quality) +
  geom_line(size = 1)

p6 <- ggplot(aes(y= WineQuality, x=total.sulfur.dioxide_mean), data = wine.by_quality) +
  geom_line(size = 1)

p7 <- ggplot(aes(y= WineQuality, x=density_mean), data = wine.by_quality) +
  geom_line(size = 1)

p8 <- ggplot(aes(y= WineQuality, x=pH_mean), data = wine.by_quality) +
  geom_line(size = 1)

p9 <- ggplot(aes(y= WineQuality, x=sulphates_mean), data = wine.by_quality) +
  geom_line(size = 1)

p10 <- ggplot(aes(y= WineQuality, x=alcohol_mean), data = wine.by_quality) +
  geom_line(size = 1)+
   ggtitle(toString(c1) )+
   theme(plot.title=element_text(size=5))

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, ncol = 4)

ggplot(aes(x=quality , y=citric.acid), data = wine)+
  geom_boxplot(fill = "grey80", colour = "#3366FF")+
  geom_jitter(fill = I('red'),shape = 21,alpha = 1/3, width = 0.15)

```


```{r, Bivariate_Plots_2, message=FALSE}
ggpairs(subset(wine, select = -c(X)),
  lower = list(continuous = wrap("points", shape = I('.'))), 
  upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

### Bivariate Analysis

Based on the ggpairs above, we observed the following relationship between the wine quality and other variables:
 - negative correlation bewteen volatile.acidity and wine quality
 - positive correlation bewteen citric.acid and wine quality
 - positive correlation bewteen sulphates and wine quality


Some interesting discovery of the relationship between the variables include:
 - fixed.acidity is positive correlated to density. This makes sense since most of the wine volume is made of water, fixed.acidity is known as more dense than water
 - fixed.acidity is negative correlated to PH. This makes sense since the more acid in a certain solution, the lower PH based on chemical knowedge.


Strongest relationship among all variables is between fixed.acidity and PH, with Corr = -0.683 
Strongest relationship for wine quality is between alcohol and wine quality, with Corr = 0.476


### Multivariate Plots Section
```{r, Multivariate_Plots}
ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()


ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()

ggplot(data = wine,
       aes(y = citric.acid, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()

ggplot(data = wine,
       aes(y = pH, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer()

```



 
### Multivariate Analysis

From the multivariate plots, we have the following discoveries:
- Fixed acidity and density are positive related to each other since the density of the acid is larger than water.  
- Critic acidity and fixed acidity also strengthened each other  
- Sulphates seems always positive related to wine quality  
- free sulfur dioxide is also positive related to total sulfur dioxide


I also created a explanatory model to model the wine quality, using the 5 factors due to their high corr with wine quality: alcohol, sulphates, volatile.acidity, citric.acid and pH.

```{r}
grid.arrange(ggplot(data = wine, aes(x = quality, y = sulphates,
                                   fill = quality)) + 
               ylab('Sulphates (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot(),
             ggplot(data = wine, aes(x = quality, y = volatile.acidity,
                                   fill = quality)) +
               ylab('Volatile Acidity (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot(), 
             ggplot(data = wine, aes(x = quality, y = citric.acid,
                                   fill = quality)) +
               ylab('Citric Acid (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot(), 
             ggplot(data = wine, aes(x = quality, y = pH,
                                   fill = quality)) +
               ylab('pH') +
               xlab('Quality') +
               geom_boxplot())
```


```{r}
ggplot(data = wine, aes(x = quality, y = alcohol,
                      fill = quality)) +
  geom_boxplot() +
  ggtitle('Alcohol Levels in Different Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')
```
 
```{r Building the Linear Model}
m1 <- lm(I(as.numeric(quality)) ~ I(alcohol), data = wine)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . +volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + pH)
mtable(m1, m2, m3, m4, m5)
```

#### Quality = 2.112 + 0.327 $\times$ alcohol + 0.673 $\times$ sulphates  -1.284 $\times$ volatile.acidity -0.297 $\times$ citric.acid -0.475 $\times$ pH


The model is in general able to reflect the relationship between wine quality and the factors listed above, and to predict the wine quality given its attributes. However, the model is based on a liner relationship, so it does not provide any information about the non-liner relationship factors such as residual sugar and sulphur dioxide.
